---
title: 'PS 211 — Discussion Worksheet #2'
subtitle: 'Fall 2026 · Frequency displays & descriptive statistics'
author: "Your Name"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

# How this worksheet works

We follow the lectures in order: first **frequency distributions and visual displays** (Lecture 3), then **central tendency** (Lecture 4). Run every chunk top-to-bottom; edit code where it says **EDIT**; type short answers at each **"Your answer:"**.

**Setup:** `pixar_data.csv` should already be in your `ps211` project folder from last time.

---

## Packages (run this once at the top)

```{r libraries, message = FALSE}
# Install once if needed (then re-comment):
# install.packages("readr")
# install.packages("dplyr")
# install.packages("ggplot2")

library(readr)
library(dplyr)
library(ggplot2)

# Load the data
pixar_data <- read_csv("pixar_data.csv")
```

---

# 1) Frequency tables

A **frequency table** answers the simplest question you can ask about a variable: *how often does each value occur?* The `count()` function builds one for you — give it the data frame and the column you want tallied.

```{r freq-table}
count(pixar_data, film_rating)   # how many films have each rating?
```

The `n` column is the **frequency** — the number of films in each category.

**Your answer:** How many films in this dataset are rated G, and how many are rated PG?

*(Type your answer here.)*

---

# 2) Grouped frequency tables

When a variable has too many distinct values to tally one at a time, we bin them into **intervals** first — a **grouped frequency table**. Release year is a good example: 23 films spread over ~25 years would give a nearly useless table, mostly one film per row. Grouping by decade fixes that.

First we make a `decade` variable, then we count it. (Don't worry about memorizing the `mutate()` line — focus on the counting.)

```{r decades}
pixar_data <- pixar_data |>
  mutate(decade = floor(release_year / 10) * 10)   # e.g., 1995 -> 1990

count(pixar_data, decade)
```

*(The `|>` symbol is the "pipe" — it means "take the thing on the left and feed it into the function on the right." You'll see it a lot in R code.)* You use pipes in R to write cleaner, more readable, and more maintainable code. The shortcut for Mac keyboard shortcut for `|>` is shift + command + M. For Windows it's Ctrl + Shift + M

**Your answer:** Which decade produced the most Pixar films in this dataset?

*(Type your answer here.)*

**Your answer:** Try running `count(pixar_data, release_year)` in the Console and compare it to the decade table above. What did grouping buy us, and what did it cost us?

*(Type your answer here.)*

---

# 3) Histograms

A **histogram** does for a numeric variable what a grouped frequency table does — it sorts values into bins — but shows the result as a picture. This is the single most useful "first look" at any numeric variable.

```{r histogram}
ggplot(pixar_data, aes(x = metacritic)) +
  geom_histogram(binwidth = 5, fill = "lightgrey", color = "black") +
  labs(title = "Metacritic Scores of Pixar Films",
       x = "Metacritic Score", y = "Number of Films") +
  theme_classic()
```

Anatomy of that `ggplot` code, line by line: the data and which column goes on x (`aes`), the type of plot (`geom_histogram`) with a bin width, the labels (`labs`), and a clean look (`theme_classic`).

**Your answer:** Describe this distribution's shape in one sentence: is it roughly symmetric, positively skewed, or negatively skewed?

*(Type your answer here.)*

**Try it:** Change `binwidth = 5` in the chunk above to `binwidth = 1`, re-run, then try `binwidth = 20`.

**Your answer:** What happens at each extreme, and why is a middle-ground bin width usually most informative?

*(Type your answer here.)*

---

# 4) Bar plots for categorical variables

Histograms are for numeric variables, where the bins sit on a continuous scale and the bars touch. For a **categorical** variable (like `film_rating`) there is nothing to bin — we just count each category and draw a **bar plot**, with gaps between the bars to show the categories are separate.

```{r barplot}
ggplot(pixar_data, aes(x = film_rating)) +
  geom_bar(fill = "lightgrey", color = "black") +
  labs(title = "Pixar Films by Rating",
       x = "Film Rating", y = "Number of Films") +
  theme_classic()
```

Notice that this bar plot shows exactly the same information as the frequency table in Section 1 — a table and a plot are two displays of the same frequency distribution.

**Your answer:** Why would it be wrong to display `film_rating` with a histogram instead?

*(Type your answer here.)*

---

# 5) Central tendency: mean, median, mode

Now we summarize a whole distribution with a single number. R has built-in functions for the mean and median:

```{r central-tendency}
mean(pixar_data$run_time)     # mean run time, in minutes
median(pixar_data$run_time)   # median run time
```

R has no built-in function for the **mode** — but you already know how to find it. The mode is just the most frequent value, so build a frequency table (Section 1!) and sort it. `arrange(desc(n))` puts the largest count on top:

```{r mode}
count(pixar_data, run_time) |>
  arrange(desc(n))
```

**Your answer:** What is the mode of `run_time`? Is it close to the mean and median?

*(Type your answer here.)*

**Try it:** Edit the chunk below to compute the **mean** and **median** of the `metacritic` column.

```{r try-central}
# EDIT THESE LINES, THEN RUN:
metacritic_mean <- 0     # REPLACE 0 with code computing the mean metacritic score
metacritic_median <- 0   # REPLACE 0 with code computing the median metacritic score

metacritic_mean
metacritic_median
```

**Your answer:** The mean is a bit *lower* than the median here. Look back at the histogram in Section 3. What does that tell you about the shape of the distribution, and why does skew pull the mean but not the median?

*(Type your answer here.)*

---

# 6) Putting it together

**Try it:** Copy the histogram code from Section 3 into the chunk below and change it to plot the distribution of **`run_time`** instead of `metacritic`. Pick a bin width you think is reasonable, and update the title and x-axis label to match.

```{r try-histogram}
# WRITE YOUR CODE HERE (copy + edit the Section 3 histogram code)

```

**Your answer:** Most Pixar films cluster around what run time? Are there any unusually long or short films (outliers)? Does the picture agree with the mean, median, and mode you computed in Section 5?

*(Type your answer here.)*

---

# ✅ Before you leave

- [ ] I built a frequency table with `count()`, and a grouped one by decade.
- [ ] I made (and modified!) a histogram and a bar plot, and I can say when to use each.
- [ ] I computed a mean, median, and mode in R.
- [ ] I knitted to HTML and showed my TF.

**Next week:** variability — how spread out a distribution is — and what happens when we take samples from a population.
